---
title: "Modeling Trade VS Natural Disasters"
author: "Nick Kent(Tuesday, May 21st 2024)"
format:
  html:
    self-contained: true
---

# Loading the data

```{r message=FALSE}
# Necessary packages
library(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)
library(GGally)
library(broom) 

# Loading in the dataset
natural_disaster <- read_csv("nat_disaster_data.csv")
country_trade_data <- read_csv("country_trade_data.csv")
```

# Data wrangling/cleaning

```{r message=FALSE}
# Creating a data set for each countries number of natural disasters per year over the last 23 years. 
nat_dis_per_year_per_country <- natural_disaster %>% 
  group_by(Country, `Start Year`) %>% 
  summarise(num_of_nat_dis = n()) %>% 
  mutate("Start Year" = as.character(`Start Year`))

# Obtain a data set for each countries trade percentage of GDP over the last 23 years. 
trade_per_year_per_country <- country_trade_data %>% 
  pivot_longer(cols = c("1999","2000","2001","2002","2003","2004","2005","2006","2007","2008","2009","2010","2011","2012","2013","2014","2015","2016","2017","2018","2019","2020","2021","2022"), names_to = "Year", values_to = "Trade share of GDP") %>% 
  select(`Country Name`, Year, `Trade share of GDP`)

# We now have a data set that contains a line for each country and year combination, and then along with that is the countries trade share of GDP and number of natural disasters in that year. 
model_data <- trade_per_year_per_country %>% 
  left_join(nat_dis_per_year_per_country, by = c("Country Name" = "Country", "Year" = "Start Year")) %>% 
  filter(!is.na(num_of_nat_dis))
```

# Creating a model

```{r message=FALSE, warning=FALSE}
model <- model_data %>%
  with(lm(`Trade share of GDP` ~ num_of_nat_dis))

model %>% 
  tidy()

model %>% 
  tidy() %>% 
  summarize(lb = estimate - 2*std.error, ub = estimate + 2*std.error)

```

# Takeaways from the model

> -   Confidence interval for the slope does not contain zero and all the values in the interval are negative which means we can be sure that there is a negative relationship between natural disasters and the trade share of GDP.
> -   The intercept tells us that if a country faces no natural disasters then on average their trade share of GDP will be about 81.77 %.
> -   The estimate for the slope tells us that on average for every one additional natural disaster that occurs in a country their trade share of GDP will drop by an average of 2.092 %.
> -   P-value for both the intercept and the slope are extremely low meaning that these estimates are statistically significant.
